\documentclass{article}

\usepackage{iclr2026_conference,times}
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

\title{Robotic Common Sense as Executable Affordance Tests}

\author{Anonymous Authors}

\begin{document}

\maketitle

\begin{abstract}
Robotic ``common sense'' is often treated as a prior over plausible facts:
mugs hold liquid, knives cut, boxes support objects.  That framing is fragile
for embodied agents because the dangerous cases are precisely those in which a
name or visual category remains familiar while the physical variables have
changed: a mug has a hole, a knife is dull, a box is wet, or the robot gripper
cannot apply the required force.  We propose Executable Affordance Test Logic
(EATL), in which a common-sense affordance predicate is not a text fact, a
static label, or a planner precondition, but a small robot-specific test program
that returns a witness or failure certificate for a given object, action, robot,
and task demand.  We give a simple lower bound showing that any label-only prior
fails on label-preserving affordance flips, and we provide a runnable synthetic
tabletop stress test covering containment, lifting, load support, cutting,
wiping, and hot carrying.  In the flip regime, a name prior reaches 0.834
accuracy with 0.134 unsafe false positives, while EATL reaches 0.972 accuracy
with 0.022 unsafe false positives at mean test cost 0.095.  The evidence is not
a real-robot result; it is an executable demonstration that the common field
assumption--object names track physical affordances--is a mechanism-level
failure point.
\end{abstract}

\section{Introduction}

Embodied agents inherit a tempting shortcut from language: if the object is
called a mug, use mug common sense.  This shortcut appears throughout robot
systems that map language, object categories, or visual detections into
affordance scores, skill choices, or planning preconditions.  It is also often
useful.  A robot should not relearn from scratch that bowls usually contain
liquid or that sponges usually wipe spills.

The problem is that common sense for robots is not only about what is likely. It
is about what is safe to do now.  A name can remain unchanged while the physical
predicate changes: a cup-shaped sieve leaks, a mug has a hairline crack, a
cardboard box collapses under a heavy object, a knife is decorative and dull, or
a plastic tray deforms under heat.  These are not exotic philosophical cases.
They are ordinary deployment failures caused by hidden state, materials, wear,
counterfeits, task demand, and embodiment.

This paper argues that robotic common sense should be operational rather than
textual.  We define a common-sense affordance claim such as
\texttt{can-contain-liquid(object, 200ml)} by the executable micro-experiment
that would make it true for this robot: pour a safe probe volume, observe
leakage and stability, and return a witness.  Similar tests can probe lift,
support, cut, wipe, or carry-hot predicates.  The central object is therefore
not a larger language model, a better dataset, or a verifier bolted onto a plan.
It is a semantics for embodied common-sense predicates.

Our contributions are:
\begin{itemize}
\item We formalize Executable Affordance Test Logic (EATL): affordance
predicates are typed test programs over robot, object, action, demand, and
observation.
\item We prove a simple impossibility result for label-only common-sense priors
under label-preserving affordance flips.
\item We release a runnable synthetic stress test showing that the broken
assumption matters under controlled physical mutations.
\item We map the novelty boundary against affordance learning, task-and-motion
planning, language-conditioned robotics, active sensing, and world models.
\end{itemize}

\section{Related Work and Novelty Boundary}

\textbf{Affordances.}
Affordances originate as action possibilities relative to an actor
\citep{gibson1979ecological}; design work distinguishes perceived affordances
and constraints \citep{norman1988psychology}.  Robotics has long treated
affordances as relations among entities, actions, and effects
\citep{sahin2007afford,montesano2008learning,stoytchev2005behavior}.
Interactive object learning and object-action complexes similarly ground
perception in action \citep{fitzpatrick2003learning,kruger2011object}.  These
works make generic affordance prediction non-novel.  Our distinction is that an
affordance predicate is accepted as robot common sense only when its executable
test has a current witness.

\textbf{Visual and learned affordance detection.}
RGB-D activity models, part-based tool affordances, and deep affordance
detectors predict likely functions from observations
\citep{koppula2013learning,myers2015affordance,do2018affordancenet}. Knowledge
representations encode object affordances for reasoning
\citep{zhu2014reasoning}.  These methods can provide candidates or priors for
EATL, but the proposed semantics rejects a prediction that lacks a
task-demand-specific test witness.

\textbf{Planning and feasibility.}
Task-and-motion planning integrates symbolic choices with geometric or
continuous feasibility checks
\citep{kaelbling2011hierarchical,toussaint2015logic,garrett2020pddlstream}.
This makes feasibility checking inside a planner non-novel.  EATL differs in
scope: it defines embodied common-sense claims before plan execution and treats
failure certificates as first-class outputs for hidden physical variables such
as holes, porosity, dull edges, and load capacity.

\textbf{Language-conditioned robotics.}
Recent systems use language models, multimodal models, or robot foundation
models to guide skill choice and manipulation
\citep{ahn2022saycan,huang2022inner,liang2023code,huang2023voxposer,driess2023palme,brohan2022rt1,brohan2023rt2,openx2023,shridhar2022cliport}.
Other work highlights that language models can struggle with planning
\citep{huang2022zeroshot,valmeekam2023plan}.  Our hostile distribution
decouples object names from causal affordance variables.  In that setting, text
priors are the object of critique, not the mechanism.

\textbf{Interactive perception, self-supervision, and world models.}
Robots can learn by poking, grasping, or predicting future observations
\citep{agrawal2016poke,pinto2016supersizing,ha2018world}.  EATL can use such
models as implementations of tests.  The new claim is narrower: the meaning of
a robot common-sense predicate is the executable evidence that certifies it for
the current robot-object-demand tuple.

\section{Executable Affordance Test Logic}

Let $o$ be an object instance, $r$ a robot embodiment, $a$ a candidate action
family, and $d$ a task demand such as volume, load, temperature, or required
precision.  A conventional prior estimates $p(y \mid \ell, x)$ where $\ell$ is
the object label and $x$ is a passive observation.  EATL instead attaches a test
program to each predicate.

\begin{definition}[Executable affordance test]
An executable affordance test for predicate $\phi$ is a tuple
$T_{\phi}=(\pi_{\phi}, g_{\phi}, c_{\phi})$.  The micro-policy $\pi_{\phi}$
chooses a bounded diagnostic action for $(r,o,a,d)$, the observation function
returns a trace $\tau$, and the guard $g_{\phi}(\tau)$ returns
\textsc{pass}, \textsc{fail}, or \textsc{abstain}.  The scalar $c_{\phi}$ is
the normalized test cost.
\end{definition}

\begin{definition}[EATL witness]
The robot may assert $\phi(r,o,a,d)$ only if it has a witness
$(T_{\phi},\tau)$ such that $g_{\phi}(\tau)=\textsc{pass}$ under the current
robot, object, action, and demand.  If the guard returns \textsc{fail}, the
trace is a failure certificate; if it returns \textsc{abstain}, the predicate is
unavailable for safe planning.
\end{definition}

Examples are deliberately mundane.  \texttt{can-contain-liquid} is tested by a
safe pour probe that observes leakage and stability.  \texttt{can-support-load}
places a calibration load below the task demand.  \texttt{can-cut-soft-food}
uses a low-force edge probe on a sacrificial material.  \texttt{can-wipe-spill}
measures absorbed volume.  \texttt{can-carry-hot} tests heat resistance and
handle availability.  These are not full task rollouts.  They are cheap
pre-task experiments whose observations target the causal variables that a name
or passive image may not expose.

\section{A Simple Lower Bound}

The formal point is intentionally small: if the only input is a label and the
deployment process can preserve that label while flipping the affordance, no
text prior can recover the hidden variable.

\begin{proposition}[Label-only error under affordance flips]
Fix an object label $\ell$ and binary affordance predicate $Y$.  Any classifier
$h$ that depends only on $\ell$ has conditional error
\[
  \Pr[h(\ell)\neq Y \mid L=\ell] \geq \min(p_{\ell},1-p_{\ell}),
\]
where $p_{\ell}=\Pr[Y=1\mid L=\ell]$.  If a deployment distribution presents
balanced same-label counterexamples, $p_{\ell}=1/2$, the error is at least
$1/2$.  If an executable test observes sufficient evidence $S$ with guard
$g(S)$ satisfying $\Pr[g(S)\neq Y]\leq \epsilon$, then the test-based predicate
has error at most $\epsilon$ on non-abstaining trials.
\end{proposition}

\textit{Proof sketch.}
For a fixed label, $h(\ell)$ is a constant.  If it predicts 1, its error is
$1-p_{\ell}$; if it predicts 0, its error is $p_{\ell}$.  The best label-only
choice therefore has error $\min(p_{\ell},1-p_{\ell})$.  The test claim follows
directly from the assumed guard error.  The result does not say EATL tests are
free or automatically available; it says the missing input channel is a real
failure mode for label-only common sense.

\section{Synthetic Stress Test}

We implement a transparent tabletop world with ten named object categories:
mug, bowl, sieve, knife, spoon, sponge, brick, cardboard box, plastic tray, and
glass.  Each instance has hidden physical variables including concavity,
porosity, holes, capacity, mass, friction, load capacity, sharpness, edge
stiffness, absorbency, heat resistance, handles, stability, and flatness.
Ground-truth affordances are deterministic functions of these variables and the
task demand.

We compare three mechanisms:
\begin{itemize}
\item \textbf{text\_prior}: a hand-coded name prior such as mugs contain liquid
and knives cut.
\item \textbf{vision\_proxy}: a passive geometry proxy that sees shape-like
variables but misses hidden material, damage, and wear.
\item \textbf{eatl}: executable tests that probe the task-specific causal
variables with small independent observation noise.
\end{itemize}

The regimes are: typical objects with mild wear; label-preserving flips where
names stay fixed while hidden physical variables change; label swaps where an
object inherits another object's physical variables but keeps its name; and a
mixed regime.  The run uses 1000 episodes and 180000 method-task rows.

\begin{table}[t]
\centering
\small
\input{results_table.tex}
\caption{Runnable synthetic stress test.  Unsafe false positives are cases in
which a method asserts an affordance that is false, e.g., approving a leaking
container or overloaded support.}
\label{tab:results}
\end{table}

Table~\ref{tab:results} shows the intended failure.  The text prior is
reasonable on typical objects, but under label-preserving flips it has 0.134
unsafe false positives.  The vision proxy has even higher unsafe false
positives because visible geometry often remains plausible while material or
damage state changes.  EATL reduces unsafe false positives to 0.022 by testing
the deciding variable.  Its mean normalized test cost is 0.095, representing
the overhead of the diagnostic micro-experiment.

\section{Discussion}

The result changes the central mechanism.  If common sense is a prior, then the
natural repair is a larger model, more data, uncertainty, or a verifier.  If
common sense is executable, the repair is different: define the predicate by
the physical evidence that would make it true for this robot and this demand.
That is why the proposed mechanism is not a new benchmark alone.  The hostile
benchmark exists to expose the assumption, but the contribution is the
operational semantics.

EATL also changes how failures are represented.  A failed pour probe is not
just a low confidence score; it is a certificate that the current object does
not satisfy the containment predicate for the demanded volume.  This is useful
for planners because the failure is local, typed, and reusable until the object
state changes.

\section{Limitations}

This paper does not show real-robot performance.  The tests are hand-specified,
the cost model is simplified, and the synthetic environment assumes the test
observes sufficient causal variables with independent noise.  Some predicates
may require destructive tests, long-horizon interaction, or specialized
sensors.  Passive perception and large multimodal models can also solve many
cases when the relevant variable is visible.  The claim is therefore not that
EATL replaces perception, planning, or learned affordance models.  The claim is
that an embodied common-sense assertion should be treated as unsupported until
the robot has executable evidence for the current physical relation.

\section{Conclusion}

Robots fail when familiar words hide changed physics.  We proposed EATL, a
semantics in which common-sense affordance predicates are certified by cheap
robot-specific tests rather than inherited from text priors.  A simple lower
bound shows why label-only priors cannot solve label-preserving affordance
flips, and a runnable synthetic stress test demonstrates the safety effect on
containment, lifting, support, cutting, wiping, and hot carrying.  The next step
is to learn or synthesize test programs and evaluate them on real robot
manipulation tasks with explicit test cost and failure risk.

\subsubsection*{Reproducibility Statement}
The repository contains the literature builder, synthetic environment, episode
logs, summary CSV, and table-generation script.  The main experiment is
reproduced with:
\begin{verbatim}
python experiments/affordance_tests.py
python scripts/write_paper_assets.py
\end{verbatim}
The literature sweep is reproduced with \texttt{python scripts/build\_literature.py};
OpenAlex rate limiting during this run was recovered using Crossref and arXiv
metadata, as recorded in \texttt{data/literature\_progress.json}.

\bibliography{references}
\bibliographystyle{iclr2026_conference}

\end{document}
